Project Protocol

### 2019-10-23 protocol.Rmd

By chance, the data plot will be detected with probability, $1/12=0.083$. This corresponds to about 2 detections out of 20 evaluatoins, and generally this is $n/12$ detections from $n$ evaluations. Let's suppose the alternative hypothesis has probability of detection $p_A (>1/12)$. If we expect the data to be detected twice as often as by chance, $p_A=1/6$. To determine the probability of observing $n/12$ detections, for one lineup, given $p_A$, we can use a binomial distribution, where $k=ceiling(n/12)$, to calculate the power:

$$P(X\geq k|p=p_A,n) = \sum_{j=k}^{n}{n \choose j} p_A^j (1-p_A)^{n-j}$$

@@

```{r echo=FALSE, eval=TRUE, fig.height=5, fig.width=6, out.width="70%"}
# Code to compute the power for given sample size and alternative p
library(tidyverse)
library(nullabor)
library(viridis)
library(knitr)
n <- 20:50
pA <- seq(1/12, 1/3, 0.002)
g <- expand.grid(n, pA)
pow <- tibble(n=g[,1], k=as.integer(ceiling(g[,1]/12)), pA=g[,2]) 
pow <- pow %>% 
  mutate(prob = pbinom(k-1, n, pA, lower.tail=FALSE)) 
ggplot(pow, aes(x=n, y=pA, fill=prob, group=pA)) + 
  geom_tile() +
  scale_fill_viridis_c("power") +
  geom_hline(yintercept=0.1666667, colour="black") +
  ylab("detect rate (pA)") + xlab("sample size (n)") +
  theme_bw()
pow %>% filter(near(pA, 0.167, 0.001)) %>% filter(prob>0.9) %>% top_n(n=10) %>% kable(digits=3)
```

The table above shows the sample sizes yielding above 90% chance of detection ($p_d$) in a single lineup assuming $p_A=1/6=0.167$ we would need approximately 50 evaluations for each lineup.

If 4 lineups are evaluated for each trend model, the chance that none produce a detection is $(1-p_d)^4$, which for $p_D>0.9$ is effectively 0.

>>>>

By chance, the data plot will be detected with probability, $1/12=0.083$. Assuming a binomial distribution we can determine the critical value that would indicate the number of detections from $n$ evaluations, corresponding to a rejection of $H_0: p=0.083$. For example, with 50 evaluations, with significance level 0.05, if there are 8 or more detections, it is better than chance, say $C$.

This critical value can be used to estimate the sample size needed, for the study, by computing the power, the probability of rejecting $H_0: 0.083$, for alternative values of $p=p_A$. If we expect the data to be detected twice as often as by chance, $p_A=1/6$. To determine the probability of observing $C$ detections from $n$ evaluations, for one lineup, to calculate the power:

$$P(X\geq C|p=p_A,n) = \sum_{j=C}^{n}{n \choose j} p_A^j (1-p_A)^{n-j}$$

@@

```{r echo=FALSE, eval=TRUE, fig.height=5, fig.width=6, out.width="70%"}
# Code to compute the power for given sample size and alternative p
library(tidyverse)
library(nullabor)
library(viridis)
library(knitr)
n <- 20:50
pA <- seq(1/12, 1/3, 0.002)
pow <- sample_size(n, 12, pA)
#g <- expand.grid(n, pA)
#pow <- tibble(n=g[,1], k=as.integer(ceiling(g[,1]/12)), pA=g[,2]) 
#pow <- pow %>% 
#  mutate(prob = pbinom(k-1, n, pA, lower.tail=FALSE)) 
ggplot(pow, aes(x=n, y=pA, fill=prob, group=pA)) + 
  geom_tile() +
  scale_fill_viridis_c("power") +
  geom_hline(yintercept=0.1666667, colour="black") +
  ylab("detect rate (pA)") + xlab("sample size (n)") +
  theme_bw()
pow %>% filter(near(pA, 0.167, 0.001)) %>% filter(prob>0.7) %>% top_n(n=10) %>% kable(digits=3)
```

The table above shows the sample sizes yielding above 70% chance of detection ($p_d$) in a single lineup assuming $p_A=1/6=0.167$ we would need approximately 50 evaluations for each lineup.

If 4 lineups are evaluated for each trend model, the chance that none produce a detection is $1-(1-p_d)^4$, which is effectively 0. With 100 participants, giving 50 evaluations per lineup, if there is an effect size twice as high as chance, then we should see it in the results.
